---
title: "Research about the most harmful type of severe weather event in U.S. in 1950-2011 using NOAA Storm Database"
output: html_document
date: "2023-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
All over the world, severe weather events have been occuring every year, and causing considerable damage to people's health and economy. We want to figure out 'What type of weather event has caused the most damage to people's health/economy in U.S. from 1950 to 2011.
As an extension of this exploration, it is expected to be able to prevent such damage by preparing precautions.
<br/>
We use the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database for our analysis. There are '913' weather events types such as 'Tornado', 'Hail', 'Lightning'. The damage to people's health is divided into 'fatality' and 'injury', and the damage to economy is divided into 'property damage' and 'crop damage' and later we make new variable 'total' that add 'property damage' and 'crop damage'. 
<br/>
As the result of our analysis, 'Tornado' is the most harmful type in both 'fatality' and 'injury' over the 61 years we've tallied, and in terms of economy, the weather type that has cost the most economic damage is 'Hurricane'.
By investigating the mechanism that these two types of event occur damage to health and economy, it is expected to be able to prepare precaution that will reduce damage from these events.

## Load necessary packages
Before analysis, We'll load some packages.<br/>
Necessary packages are 'dplyr', 'stringr', 'ggplot2'.<br/>
We'll load some additional packages that are dependent.

```{r, message=FALSE}
library(base)
library(stats)
library(dplyr)
library(stringr)
library(ggplot2)
```


## Data Processing
In this part, We will read raw data('repdata_data_StormData.csv.bz2') and process it to make our future analysis easier.
<br/><br/>
First, Let's read raw data.

```{r, cache=TRUE}
Storm <- read.csv('repdata_data_StormData.csv.bz2')
```


And let's take a quick look at the data!
```{r}
dim(Storm)
names(Storm)
head(Storm)
```

There are `r dim(Storm)[1]` rows and `r dim(Storm)[2]` columns.
And this data contain information about severe weather events(like Hurricane, Flood..) including the event's location, damage it caused.<br/>

So, There are kinds of columns like 'event type'(EVTYPE), 'state where the events occured'(STATE), 'when they occured'(BGN_DATE, BGN_TIME, END_DATE, END_TIME).<br/>

Among this columns, We will extract only 7 columns we need for our analysis.
<br/>
These columns are "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP".

```{r, cache=TRUE}
needed_columns <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG",
                    "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
selected_Storm <- Storm %>% select(all_of(needed_columns))
```

```{r}
dim(selected_Storm)
names(selected_Storm)
```

So, selection of columns we want is completed.

Also, there are some inappropriate EVTYPE in this dataset.
There are 2 cases about it.

1. name containing blank space like "   HIGH SURF ADVISORY".
2. name that doesn't mean any type of event like "Summary August 2-3"(containing 'summary')

So, we will trim corresponding rows in first case, and in second case, remove corresponding rows.

```{r, cache=TRUE}
selected_Storm$EVTYPE <- str_trim(selected_Storm$EVTYPE)
good <- !grepl("Summary", selected_Storm$EVTYPE)
Storm3 <- selected_Storm[good, ]
```
So, in our first question(Across the United States, which types of events are most harmful with respect to population health?), we will use this processed 'Storm3' data.
<br/><br/>


And for our second Question(Across the United States, which types of events have the greatest economic consequences?), we need some additional processing procedure.
<br/><br/>
In 'Storm3' data, There are 2 variables("PROPDMGEXP", "CROPDMGEXP"). This 2 variables mean the Unit of the number of property damage("PROPDMG"), crop damage("CROPDMG").(in dollar amount).
<br/>
As the example value of this 2 variables("PROPDMGEXP", "CROPDMGEXP"), "K"(Kilo) means 10^3, "M"(Million) means 10^6, "b"(Billion) means 10^9. And the "" Unit is used when the "PROPDMG", "CROPDMG" value is zero.
<br/><br/>
Other than these 4 representative values of Unit, there are also values like "m", "+", "5", "?", "h", "7", "H", "-". Because we could not able to figure out meaning of these variables, We will remove rows that have value like them.
<br/>
And we will also remove the rows that both "PROPDMG", "CROPDMG" value are zero. Let's do it.

```{r, cache=TRUE}
filter_Storm <- Storm3 %>% 
  filter(PROPDMG != 0 | PROPDMG != 0) %>%
  filter(PROPDMGEXP %in% c("K", "M", "B", "")) %>%
  filter(CROPDMGEXP %in% c("K", "M", "B", ""))
```
```{r}
unique(filter_Storm$PROPDMGEXP)
unique(filter_Storm$CROPDMGEXP)
```

So, Only the rows that has one of the Unit value among ("", "K", "M", "B") are left.
<br/><br/>
In our second question, we will use this 'filter_Storm' data.

## Question 1. Across the United States, which types of events are most harmful with respect to population health?
Specifically, which types of events have caused the highest number of fatalities and injuries?
<br/>
For answering it, we will group our processed data(Storm3) by 'EVTYPE' variable and calculate sum of the fatality and injury for each group.
```{r, cache=TRUE}
group_Storm <- Storm3 %>% group_by(EVTYPE) %>% 
  summarize(fatality = sum(FATALITIES), injury = sum(INJURIES))
```
and arrange the result in order of the fatalities and injuries.
```{r}
ordering1 <-
  group_Storm[order(group_Storm$fatality, decreasing = TRUE),]

ordering2 <-
  group_Storm[order(group_Storm$injury, decreasing = TRUE),]
```

The result about this Question 1 is suggested in 'Result' part.

## Question 2. Across the United States, which types of events have the greatest economic consequences?
In our raw dataset(Storm), we can identify the economic consequences through 2 variables.("PROPDMG"(property damage), "CROPDMG"(crop damage)). And we will calculate the total economic damage caused by specific event by adding these 2 variables. For this, we will make 2 new colomns that contain the damage amount in a unified unit.(each name is "property" and "crop"). And after it, we will make an additional column that means total amount of damage in dollar by adding 2 columns("property", "crop").
<br/><br/>
Let's do it!

```{r, cache=TRUE}
for(i in 1:nrow(filter_Storm)){
  if(filter_Storm$PROPDMGEXP[i] == ""){
    filter_Storm$property[i] <- 0
  }else if(filter_Storm$PROPDMGEXP[i] == "K"){
    filter_Storm$property[i] <- filter_Storm$PROPDMG[i] * 1000
  }else if(filter_Storm$PROPDMGEXP[i] == "M"){
    filter_Storm$property[i] <- filter_Storm$PROPDMG[i] * 1000000
  }else if(filter_Storm$PROPDMGEXP[i] == "B"){
    filter_Storm$property[i] <- filter_Storm$PROPDMG[i] * 1000000000
  }
}
```
First make 'property' column by combining 'PROPDMG' column with 'PROPDMGEXP'.
And also do the same thing with 'CROPDMG' and 'CROPDMGEXP' to make 'crop' column.

```{r, cache=TRUE}
for(i in 1:nrow(filter_Storm)){
  if(filter_Storm$CROPDMGEXP[i] == ""){
    filter_Storm$crop[i] <- 0
  }else if(filter_Storm$CROPDMGEXP[i] == "K"){
    filter_Storm$crop[i] <- filter_Storm$CROPDMG[i] * 1000
  }else if(filter_Storm$PROPDMGEXP[i] == "M"){
    filter_Storm$crop[i] <- filter_Storm$CROPDMG[i] * 1000000
  }else if(filter_Storm$PROPDMGEXP[i] == "B"){
    filter_Storm$crop[i] <- filter_Storm$CROPDMG[i] * 1000000000
  }
}
```
```{r}
options(scipen = 999)
```
```{r}
head(unique(filter_Storm$property))
head(unique(filter_Storm$crop))
```
So, 2 new columns has made.
<br/><br/>
And let's make 'total' column by adding 'property' and 'crop' columns.

```{r}
filter_Storm$total <- filter_Storm$property + filter_Storm$crop
```
```{r}
head(unique(filter_Storm$total))
```
Okay, the new column we could use to calculate the amount of damage has made.
<br/>
So, which types of events have the greatest economic consequences?
<br/>
For answering it, we will group 'filter_Storm' data by 'EVTYPE' variable and calculate sum of 'total' column for each group.

```{r, cache=TRUE}
final_Storm <- filter_Storm %>% group_by(EVTYPE) %>%
  summarize(economic = sum(total))
```
And arrange it in order of 'economic'.

```{r}
ordering3 <- final_Storm[order(final_Storm$economic, decreasing = TRUE),]
```

The result about this Question 2 is also suggested in 'Result' part.

## Results

### Question 1. Across the United States, which types of events are most harmful with respect to population health?
Let's see the result about the first question.

```{r}
##Plot it
par(mfrow = c(1,2), cex.axis=0.5, cex.main=.75)
barplot(ordering1$fatality[1:3], names.arg = ordering1$EVTYPE[1:3],
        xlab = "Type", ylab = "Fatality")
title(main = "Top3 Sum of fatalities by types of event 1950-2011")
barplot(ordering2$injury[1:3], names.arg = ordering2$EVTYPE[1:3],
        xlab = "Type", ylab = "Injury")

title(main = "Top3 Sum of Injuries by types of event 1950-2011")
```




Among various types of events, Tornado has caused the most fatalities and injuries.
<br/>
It caused `r ordering1[1,2]` fatalities and `r ordering2[1,3]` injuries in 1950-2011.


### Question 2. Across the United States, which types of events have the greatest economic consequences?

Let's see the result about question 2

```{r}
g <- ggplot(ordering3[1:3,], aes(reorder(EVTYPE, economic, decreasing=TRUE), economic/10^9))
g + geom_bar(stat='identity') + labs(x='Type', y='Economic Damage(Billon dollars)',
                      title = 'Top3 Type of Event that caused Economic
                      Damage 1950-2011') + 
  theme(plot.title = element_text(hjust = 0.5))
```

So, the type that caused the most economic damage across United States in 1950-2011 is,
'Hurricane'. It cost almost `r ordering3[1,2]/10^9` Billion dollars!

